Predictions From Day 0 Community Structures

### 2016-08-08 code/HumanCdGF_microbiota_predict.Rmd
Add log transformed cfu

#read in files
meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 'group')
shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header = T, row.names='Group')

#### main subset of meta file with all mouse and day 0 info
day_0_meta <- meta_file[which(meta_file$day==0),!is.na(meta_file['430-307-D0',])]
day_0_meta$RowNames <- rownames(day_0_meta)
day_0_meta$Euth_Early <-ifelse(day_0_meta$Euthanized %in% 10, 'No','Yes')
####
#    function to move mouse data from rows by mouse and days --> rows by mouse and columns by days
####
# select data to move
cdiff_effect <- c("cdiff_cfu","percent_weightLoss.x","Log_repiricoal_dilution")
# function input meta_file as DATAFRAME and mouse dataframe as INPUT
combine_days_by_mouse <- function(DATAFRAME, INPUT){
          OUTPUT <- INPUT
     for(DAY in 1:10){
          NEW_ROW <- DATAFRAME[which(DATAFRAME$day==DAY),c(cdiff_effect, 'Mouse_ID')]
          names(NEW_ROW)[!names(NEW_ROW) %in% 'Mouse_ID'] <- 
               paste('Day_',paste(DAY,paste('_',names(NEW_ROW)[!names(NEW_ROW) %in% 'Mouse_ID'], sep=''), sep=''),sep='')
          OUTPUT <- merge(OUTPUT, NEW_ROW, by='Mouse_ID', all.x=TRUE)
     }
     return(OUTPUT)
}
rf_meta_df <- combine_days_by_mouse(meta_file[!meta_file$sample_type=='cecum',], day_0_meta)

# remove all columns with no data - toxin missing from some days
rf_meta_df <- rf_meta_df[,colSums(is.na(rf_meta_df)) != nrow(rf_meta_df)]
# add columns for data from day euthanized
day_euthanized_cdiff <- meta_file[which(meta_file$sample_type=='stool' & meta_file$day == meta_file$Euthanized), c(cdiff_effect,'Mouse_ID')]
names(day_euthanized_cdiff)[!names(day_euthanized_cdiff) %in% 'Mouse_ID'] <- 
     paste('Day_Euth_',names(day_euthanized_cdiff)[!names(day_euthanized_cdiff) %in% 'Mouse_ID'], sep='')
rf_meta_df <- merge(rf_meta_df,day_euthanized_cdiff, by='Mouse_ID', all.x=TRUE)

#### subset shared
#Create df with relative abundances
day_0_shared <- shared_file[rownames(day_0_meta),]
rel_abund_day_0 <- 100*day_0_shared/unique(apply(day_0_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_day_0_01 <- apply(rel_abund_day_0, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_day_0 <- rel_abund_day_0[,OTUs_day_0_01]
OTU_list <- names(rel_abund_day_0)

#Merge OTU data with mouse/exp data
Predict_df <- merge(rf_meta_df, rel_abund_day_0, by.x='RowNames', by.y='row.names')
Predict_df <- Predict_df[,!names(Predict_df) %in% c('Mouse_ID','RowNames','cage_id')]

>>>>

# read in files
meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 'group')
shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header = T, row.names='Group')

# add log transformed CFU
meta_file$log_cfu <- log10(meta_file$cdiff_cfu + 1)

# remove cecum samples
meta_file <- meta_file[! meta_file$sample_type == 'cecum', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
# use %in% instead of == because to ignore NA
# rename day 0 communities by mouse id
rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% rownames(stool_shared) ]
rel_abund_stool <- 100*stool_shared/unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[,OTUs_stool_0_01]
OTU_list <- colnames(rel_abund_stool)

#Merge OTU data with mouse/exp data
#Predict_df <- merge(rf_meta_df, rel_abund_stool, by.x='RowNames', by.y='row.names')
#Predict_df <- Predict_df[,!names(Predict_df) %in% c('Mouse_ID','RowNames','cage_id')]

#predict day cfu by day 0 community
predict_df <- merge(meta_file[meta_file$day==2, c('Mouse_ID', 'log_cfu') ], stool_shared, by='Mouse_ID', by.y='row.names', all.y = T)
predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]

RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu),], importance = TRUE)

varImpPlot(RF)
ggplot(data = data.frame(actual=predict_df[,'log_cfu'],pred=predict(RF,predict_df)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
percent_var_expl <- RF$rsq[length(RF$rsq)]*100
